
---
title: "Perth multiple station analysis"
output:
  html_document:
    toc: yes
    toc_depth: '2'
    toc_float:
      collapsed: no
    theme: darkly
    highlight: tango
    df_print: paged
    code_folding: hide
---

Code setup:
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## import and clean data

First we import our data including the Perth Airport Station.

I have downloaded data for several weather stations in the greater Perth area. These were downloaded in two batches so they are in two seperate CSVs. Thankfully, both are in a tidy format by default which makes them quite easy to wrangle. 

One batch is American units so we have to convert those to metric so they fit with the rest of the data:

```{r}
suppressMessages(library(data.table))
suppressMessages(library(tidyverse))
suppressMessages(library(ggdark))
suppressMessages(library(patchwork))
suppressMessages(library(weathermetrics))
PER_airport <- fread("../data/3184544.csv")
#filter for usability
#turns out the data attributes are useless lol
PER_airport <- PER_airport %>% select(!ends_with("_ATTRIBUTES"))

NOAA_other_stations_1_freedomUnits <- fread("../data/3186152.csv")
NOAA_other_stations_1 <- NOAA_other_stations_1_freedomUnits %>% 
  #drop extra cols
  select(!c("DAPR","DATN","DATX","DWPR","MDPR","MDTN","MDTX","TAVG")) %>% 
  mutate(
    PRCP = PRCP*25.4,
    TMAX = fahrenheit.to.celsius(TMAX),
    TMIN = fahrenheit.to.celsius(TMIN)
  )
NOAA_other_stations_2 <- fread("../data/3186197.csv") %>% 
  select(!c("DAPR","DATN","DATX","DWPR","MDPR","MDTN","MDTX"))

NOAA_other_stations_all <- rbind(NOAA_other_stations_1,NOAA_other_stations_2)
PER_airport <- PER_airport %>% select(!c("TAVG","SNWD"))
NOAA_other_stations_all <- rbind(NOAA_other_stations_all, PER_airport)
```

Add some convenience columns to the dataset, and sanity check rainfall between stations; they all look similar which is what we are expecting here.
```{r}
NOAA_other_stations_all <- NOAA_other_stations_all %>% 
  mutate(
    Year = format(DATE, "%Y") %>% as.numeric(),
    Month = month(DATE),
    Day = format(DATE, "%d") %>% as.numeric(),
    #bool for any rain
    did_rain = PRCP > 0,
    Month_name = month.name[Month] %>% fct_relevel(month.name)
  )

NOAA_other_stations_all %>% 
  group_by(NAME) %>% 
  summarise(
    mean_rainfall = mean(PRCP, na.rm = T),
    mean_TMIN = mean(TMIN, na.rm = T),
    mean_TMAX = mean(TMAX, na.rm = T),
  )
```

## data completeness

The data is incomplete in a number of ways:
1. While the full set of dates runs between 1901-2022. No station has data for the entire period
2. Stations lack some data for periods when they were active.

```{r}
NOAA_other_stations_all %>% summary
```

### periods with recorded data for each station

We want to know coverage for rainfall and temperature for each station.

```{r}
NOAA_other_stations_all %>% 
  group_by(NAME) %>% 
  summarise(
    minDate = min(DATE),
    maxDate = max(DATE),
  )
```


