
---
title: "Perth multiple station analysis"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    toc_float:
      collapsed: no
    theme: darkly
    highlight: tango
    df_print: paged
    code_folding: hide
---

Code setup:
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## import and clean data

First we import our data including the Perth Airport Station.

I have downloaded data for several weather stations in the greater Perth area. These were downloaded in two batches so they are in two seperate CSVs. Thankfully, both are in a tidy format by default which makes them quite easy to wrangle. 

One batch is American units so we have to convert those to metric so they fit with the rest of the data:

```{r}
suppressMessages(library(data.table))
suppressMessages(library(tidyverse))
suppressMessages(library(ggdark))
suppressMessages(library(patchwork))
suppressMessages(library(weathermetrics))
PER_airport <- fread("../data/3184544.csv")
#filter for usability
#turns out the data attributes are useless lol
PER_airport <- PER_airport %>% select(!ends_with("_ATTRIBUTES"))

NOAA_other_stations_1_freedomUnits <- fread("../data/3186152.csv")
NOAA_other_stations_1 <- NOAA_other_stations_1_freedomUnits %>% 
  #drop extra cols
  select(!c("DAPR","DATN","DATX","DWPR","MDPR","MDTN","MDTX","TAVG")) %>% 
  mutate(
    PRCP = PRCP*25.4,
    TMAX = fahrenheit.to.celsius(TMAX),
    TMIN = fahrenheit.to.celsius(TMIN)
  )
NOAA_other_stations_2 <- fread("../data/3186197.csv") %>% 
  select(!c("DAPR","DATN","DATX","DWPR","MDPR","MDTN","MDTX"))

NOAA_all_stations <- rbind(NOAA_other_stations_1,NOAA_other_stations_2)
PER_airport <- PER_airport %>% select(!c("TAVG","SNWD"))
NOAA_all_stations <- rbind(NOAA_all_stations, PER_airport)
```

Add some convenience columns to the dataset, and sanity check rainfall between stations; they all look similar which is what we are expecting here.
```{r}
NOAA_all_stations <- NOAA_all_stations %>% 
  mutate(
    Year = format(DATE, "%Y") %>% as.numeric(),
    Month = month(DATE),
    Day = format(DATE, "%d") %>% as.numeric(),
    #bool for any rain
    did_rain = PRCP > 0,
    Month_name = month.name[Month] %>% fct_relevel(month.name)
  )

NOAA_all_stations %>% 
  group_by(NAME) %>% 
  summarise(
    mean_rainfall = mean(PRCP, na.rm = T),
    mean_TMIN = mean(TMIN, na.rm = T),
    mean_TMAX = mean(TMAX, na.rm = T),
  )
```

## data completeness

The data is incomplete in a number of ways:
1. While the full set of dates runs between 1901-2022. No station has data for the entire period
2. Stations lack some data for periods when they were active.

We can explore this a bit with R. First, we are interested in missing data in temperature and rainfall data.

There are only a few thousand missing rainfall records, but 84754 rows of missing data for TMIN and TMAX. This is `r (84754/264657)*100`% of the rows! This is because there are large gaps in temperature records while rainfall records have continued for the same period.

```{r}
NOAA_all_stations %>% dim
NOAA_all_stations %>% count(is.na(TMAX))
NOAA_all_stations %>% count(is.na(TMIN))
NOAA_all_stations %>% count(is.na(PRCP))
```

We can inspect the percent missing per station. Some stations are missing over 80% of records, while several stations have very complete TMAX records while rainfall was recorded.

```{r}
NOAA_all_stations %>% 
  group_by(NAME) %>% 
  count(isna_TMAX=!is.na(TMAX)) %>% 
  pivot_wider(NAME, values_from = n, names_from = isna_TMAX) %>% 
  mutate(percentage_missing = (`FALSE`/(`FALSE`+`TRUE`))*100) %>% 
  arrange(desc(percentage_missing))
```


### periods with recorded data for each station

We want to know coverage for rainfall and temperature for each station.

What about specific for each kind of data?

For where data is available, we have complete records of rainfall. But temperature has large gaps for several stations.

```{r, fig.width=12, fig.height=12}
dates_covered_by_stations <- NOAA_all_stations %>% 
  group_by(NAME) %>% 
  summarise(
    minDate = min(DATE),
    maxDate = max(DATE),
  )
dates_covered <- dates_covered_by_stations %>% 
  ggplot(aes(NAME, ymin=minDate, ymax=maxDate)) +
  geom_linerange() +
  dark_theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  xlab("Station Name") + ylab("Date range (Years)") +
  ggtitle("Date range available for each station")

dates_covered_by_variable <- NOAA_all_stations %>% 
  pivot_longer(c(TMAX,TMIN,PRCP), names_to = "measurement") %>%
  # filter(NAME == "ROTTNEST ISLAND LIGHTHOUSE, AS") %>% 
  ggplot(aes(measurement, DATE, colour=!is.na(value))) +
  geom_point() +
  dark_theme_bw() +
  facet_wrap(vars(NAME)) +
  ggtitle("Dates where values are available for each station")

dates_covered + dates_covered_by_variable
```

We can do some quick checks using the non-filtered data. But we will need to use filtered data to ensure robust modelling later.

```{r, fig.width=10}
yearly_statistics <- NOAA_all_stations %>% 
  group_by(NAME, Year) %>% 
  summarise(
    meanRainfall = mean(PRCP, na.rm = T),
    totalRainfall = sum(PRCP, na.rm = T),
    meanMaximumTemperature = mean(TMAX),
    meanMinimumTemperature = mean(TMIN),
  ) 

yearly_statistics %>% 
  ggplot(aes(Year, totalRainfall, colour = NAME)) +
  geom_point(alpha=0.5) +
  geom_smooth(method="lm", se = F) +
  dark_theme_bw()
```

We can see that the two stations on Rottnest are showing roughly similar trends but because of the way the data is split the regression trends predict opposite directions of effect for total rainfall. I think it would be best to include these as a single dataset and then use the station name as a factor variable to account for the difference between each station.

```{r, fig.width=10}
yearly_statistics %>% 
  filter(str_detect(NAME, "ROTTNEST")) %>% 
  ggplot(aes(Year, totalRainfall, colour = NAME)) +
  geom_point(alpha=0.5) +
  geom_smooth(method="lm", se = F) +
  dark_theme_bw() +
  ggtitle("total rainfall each year for Rottnest Island from both stations")
Rotto_yearly_statistics <- NOAA_all_stations %>% 
  group_by(NAME, Year) %>%
  summarise(
    totalRainfall = sum(PRCP, na.rm = F),
    naRainfallCount = sum(is.na(PRCP)),
  ) %>% 
  filter(str_detect(NAME, "ROTTNEST"), Year >= 1907)

Rotto_yearly_statistics_fewNa <- Rotto_yearly_statistics %>% 
  filter(naRainfallCount < 10)

Rotto_yearly_statistics %>% 
  ggplot(aes(Year, totalRainfall)) +
  geom_point() +
  geom_smooth(method="lm") +
  dark_theme_bw()
```

After accounting for the station name as an interaction term we can see that the resulting 

```{r}
Rotto_yearly_statistics %>% 
  ggplot(aes(Year, totalRainfall, colour = NAME)) +
  geom_point(alpha=0.5) +
  geom_smooth(method="lm", se = F) +
  dark_theme_bw() +
  ggtitle("total rainfall each year for Rottnest Island from both stations")

model_rotto_rainfall_1 <- lm(totalRainfall ~ Year * NAME, data=Rotto_yearly_statistics_fewNa)
summary(model_rotto_rainfall_1)
```






