
# comparison of summer 2023 with previous years

---
title: "Perth rainfall analysis"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    toc_float:
      collapsed: no
    theme: darkly
    highlight: tango
    df_print: paged
    code_folding: hide
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## import and clean data

Import data and add some useful columns to our rain dataframe. This includes a boolean for whether it rained. Really just allows me to have more comprehensible code and saves me adding an extra label to my ggplots.

```{r}
#import libraries we want
suppressMessages(library(data.table))
suppressMessages(library(tidyverse))
suppressMessages(library(ggdark))
suppressMessages(library(patchwork))
#read data
NOAA_data <- fread("../data/3184544.csv")
#filter for usability
#turns out the data attributes are useless lol
NOAA_data <- NOAA_data %>% select(!ends_with("_ATTRIBUTES"))
#add month and day cols for easier group-by summaries later
NOAA_data <- NOAA_data %>%
    mutate(
        Year = format(DATE, "%Y") %>% as.numeric(),
        Month = month(DATE),
        Day = format(DATE, "%d") %>% as.numeric(),
        #bool for any rain
        did_rain = PRCP > 0,
        Month_name = month.name[Month] %>% fct_relevel(month.name)
    )
NOAA_data %>% head(2)
```

Read in BOM data for jan and feb 2023 (as of 14th Feb 2023):

The downloaded csv files contain unusual byte codes that are breaking `fread` and `read.csv`. They are in the header, so I can bypass these by manually adding the header in.

```{r}
BOMheaderCols <- c(
    "Date",
    "Minimum temperature (C)",
    "Maximum temperature (C)",
    "Rainfall (mm)",
    "Evaporation (mm)",
    "Sunshine (hours)",
    "Direction of maximum wind gust ",
    "Speed of maximum wind gust (km/h)",
    "Time of maximum wind gust",
    "9am Temperature (C)",
    "9am relative humidity (%)",
    "9am cloud amount (oktas)",
    "9am wind direction",
    "9am wind speed (km/h)",
    "9am MSL pressure (hPa)",
    "3pm Temperature (C)",
    "3pm relative humidity (%)",
    "3pm cloud amount (oktas)",
    "3pm wind direction",
    "3pm wind speed (km/h)",
    "3pm MSL pressure (hPa)"
)
#read in
dec2022 <-
    fread("../data/IDCJDW6110.202212.csv",
          encoding = "UTF-8",
          skip = 7)
jan2023 <-
    fread("../data/IDCJDW6110.202301.csv",
          encoding = "UTF-8",
          skip = 7)
feb2023 <-
    fread("../data/IDCJDW6110.202302.csv",
          encoding = "UTF-8",
          skip = 7)
#drop extra column
dec2022 <- dec2022 %>% select(!"V1")
jan2023 <- jan2023 %>% select(!"V1")
feb2023 <- feb2023 %>% select(!"V1")
colnames(dec2022) <- BOMheaderCols
colnames(jan2023) <- BOMheaderCols
colnames(feb2023) <- BOMheaderCols
#combine
summer2023 <- rbind(dec2022, jan2023, feb2023)
#rename temp and rainfall cols to match NOAA
summer2023 <- summer2023 %>%
    drop_na() %>%
    rename(TMIN = `Minimum temperature (C)`,
           TMAX = `Maximum temperature (C)`,
           PRCP = `Rainfall (mm)`) %>%
    #add in convenience columns
    mutate(
        Year = format(Date, "%Y") %>% as.numeric(),
        Month = month(Date),
        Day = format(Date, "%d") %>% as.numeric(),
        #bool for any rain
        did_rain = PRCP > 0,
        Month_name = month.name[Month] %>% fct_relevel(month.name),
        #we can use row_number to get "days since dec 1st" which we call "daysOfSummer"
        daysOfSummer = row_number()
    )
summer2023 %>% head(2)
```

```{r, fig.width=9, fig.height=5}
#get summer 2022 data from NOAA table
summer2022 <- NOAA_data %>% 
    filter(
        (Year == 2022 & (Month == 1 | Month == 2)) | (Year == 2021 & Month == 12)
    ) %>% 
    mutate(daysOfSummer = row_number())
#combine the two summers because it is easier...
#add identifer tag to each summer
summer2022$tag <- "summer2022"
summer2023$tag <- "summer2023"
#select and rbind
summersDF <- rbind(
    summer2022 %>% select(tag, daysOfSummer, Year, Month, Day, Month_name, PRCP, TMIN, TMAX),
    summer2023 %>% select(tag, daysOfSummer, Year, Month, Day, Month_name, PRCP, TMIN, TMAX)
)
scatterSummer <- summersDF %>% 
    ggplot(aes(daysOfSummer, TMAX, colour = tag)) +
    geom_point(alpha=0.5) +
    geom_smooth(se=F) +
    # ggtitle("Maximum daily temperature at Perth Airport\nin summer '22 and '23") +
    theme_classic() +
    theme(legend.position="none") +
    xlab("days since dec 1st")
set.seed(9001)
boxplotSummer <- summersDF %>% 
    ggplot(aes(tag, TMAX, fill = tag)) +
    geom_jitter(width = 0.1, aes(colour = tag)) +
    geom_boxplot() +
    theme_classic() +
    xlab("") + ylab("")
summerFig <- scatterSummer + boxplotSummer
summerFig + plot_annotation(
        title = "Maximum daily temperature at Perth Airport in summer '22 and '23"
    )
ggsave("../images/summer23vs22.png")
```

```{r}
#hist
summersDF %>% 
    filter(Month == 1) %>% 
    ggplot(aes(TMAX, colour = tag)) +
    geom_density()
```
```{r}

```



Last two summers vs all past summers

```{r}
ghostsOfSummersPast <- NOAA_data %>% 
    filter(Year < 2021) %>% 
    filter(Month %in% c(1,2,12))
#add identifer tag to each summer
ghostsOfSummersPast$tag <- "allPreviousSummers"
ghostsOfSummersPast <- ghostsOfSummersPast %>% 
    mutate(daysOfSummer = row_number()) %>% 
    select(tag, daysOfSummer, Year, Month, Day, Month_name, PRCP, TMIN, TMAX)
#select and rbind
summersDF <- rbind(
    summersDF,
    ghostsOfSummersPast
)
set.seed(9001)
summersDF %>% 
    ggplot(aes(tag, TMAX, fill = tag)) +
    geom_jitter(width = 0.1, alpha=0.2, aes(colour = tag)) +
    geom_boxplot() +
    theme_classic() +
    ggtitle("Distribution of maximum daily temperature for summer at Perth Airport")
```



